Sensor Clusterization in D-optimal Design in Infinite Dimensional Bayesian Linear Inverse Problems

### 2024-08-25 latex/intro.tex
answered refs

%% REV 1, 2
Clusterization should not be confused with replication. Replication
requires that the experimentalist executes multiple trials under
circumstances that are \emph{nominally identical} \cite[Section
  1.2.4]{morris2011}. Replication is commonly viewed as a beneficial
and even necessary aspect of optimal experimental design
\cite{fisher1949design, morris2011, schafer2001replication}. Similarly
to a design implementing replication, a clustered design reduces the
signal-to-noise ratio of the repeated measurements
\cite{telford2007brief}. The difference is that a clustered design
takes repeated measurements at the expense of other quantities not
measured at all.

%% Rev 2
For example, \cite{fisher1949design}, suggested repeating his famous
milk and tea experiment in order "to be able to demonstrate the
predominance of correct classifications in spite of occasional
errors". Unfortunately, in the experiments we consider, replication is
impossible. For example, in the MRI problem, we cannot generate an
individual nominally identical to the one we wish to scan.

%% Rev 2
To further illustrate the difference between replication and
clusterization, consider an experiment measuring the effect of
rainfall on grass growth \cite{fay2000rainfall}. The experiment
involved four rainfall manipulation "treatments" (i.e.~simulating
different timing and quantity of rainfall), each replicated three
times over a different plot of land. Indeed, it seems reasonable for
researchers to replicate the phenomenon they are trying to study. A
clustered design in such an experiment would imply the researchers
should take a repeated measurement \emph{on the same plot}, at the
expense of measuring grass growth in other plots. To conclude our
short discussion on replication vs.~clusterization: these are
fundamentally different concepts and even though replication is quite
intuitive, it is inapplicable to the inverse problems we consider in
this manuscript.

@@

%% REV 1
%% REV 2
When clusterization arises, we believe it should serve as a warning
signal to practitioners. In the inverse problem of the 1D heat
equation, clusterization occurs primarily because Laplacian
eigenvectors do not decay in $\Omega$. Consequently, measuring $u(x_1,
T)$ at some point $x_1 \in \Omega$ provides information about
$u(x_2,T)$ for distant points $x_2 \in \Omega$. Intuitively, this
should not happen: for small $T$, the heat distribution at $x_1$
should give very little knowledge on the heat distribution at
$x_2$. This behavior stems from a well-known property of the heat
equation: it allows information to spread instantly across the
computational domain \cite{renardy2006PDE}. In reality, heat (and
information) propagate at finite speeds. Of course, the known physical
barrier for information spread is the speed of light, but we expect
heat and information to spread considerably slower. Thus,
clusterization may suggest that our physical model itself is at fault.

@@

%% REV 1
%% REV 2
Our choice of prior is perhaps an even greater contributor to
clusterization. Our choice of Gaussian prior similarly implies
information is shared between distant locations in $\Omega$. Thus, we
suggest refraining from choosing Gaussian priors with inverse
Laplacian covariance operators. Rather, non-Gaussian priors should be
employed instead \cite{hosseini2017, hosseini2019}. So, clusterization
may suggest we give more attention to the choice of prior.

%% REV 1
Given the above discussion, it is our belief that the methods
suggested by other authors to avoid clusterization are merely
overlooking the problem. We do not view clustered designs as
inherently bad on their own. Rather, we suggest that if a clustered
design arises, then a practitioner should revisit their probabilistic
and physical modeling assumptions. If the practitioner is confident in
their assumptions, then clustered designs should be avoided only to
the extent necessitated by the physical measuring apparatus.

Another For example, \cite{tekentrup2020} provides convergence analysis for
various designs with differing space filling properties. They showed
that convergence rate depends on how these designs fill space. We
expect better convergence rates for D-optimal designs, even though
D-optimal designs may demonstrate clusterization. The reason is that
D-optimal designs explore space in a way tailored to the problem
studied. Consequently, D-optimal designs cluster because they first
aim to measure what matters most; see the discussion following
Theorem \ref{thm:char} for details.

@@

We expect D-optimal designs to judicious exploration of
design space, as implemented by a D-optimal design to give better
convergence compared to randomly sampling the design
space \cite{knapik2011}.

%% Clusterization is a peculiar phenomenon and it is perfectly reasonable
%% for someone to argue against the D-optimality criterion based on the
%% fact that it results in clustered designs. We have seen, however that
%% there is a perfectly reasonable explanation for clusterization. We
%% have shown that clusterization is an inevitable consequence of having
%% a problem with some modes where uncertainty decays faster than others.

\subsection{Limitations}\label{subsec:limitations}
The main limitation of this study is that our generic model does not
correspond to any specific real-life problem. Specifically, in its
current form, our model does not allow point evaluations. Thus, while
our model is generic enough to be analytically tractable, some may
argue that our model is too far removed from any real application. To
these claims I would answer that scientists have a long history of
studying models that are bare-bones simplifications of real systems,
e.g.~the Ising model \cite{cipra1987}, the Lorenz system \cite{brin},
the Lotka-Volterra equations \cite{logan2006}, the Carnot engine
\cite{kardar2007} and many others.

>>>>

%% Replication 1
Clusterization should not be confused with replication. Replication
requires that the experimentalist executes multiple trials under
circumstances that are \emph{nominally identical} \cite[Section
  1.2.4]{morris2011}. Replication is commonly viewed as a beneficial
and even necessary aspect of optimal experimental design
\cite{fisher1949design, morris2011, schafer2001replication}. Similarly
to a design implementing replication, a clustered design reduces the
signal-to-noise ratio of the repeated measurements
\cite{telford2007brief}. The difference is that a clustered design
takes repeated measurements at the expense of other quantities not
measured at all.

%% Replication 2
For example, \cite{fisher1949design}, suggested repeating his famous
milk and tea experiment in order "to be able to demonstrate the
predominance of correct classifications in spite of occasional
errors". Unfortunately, in the experiments we consider, replication is
impossible. For example, in the MRI problem, we cannot generate an
individual nominally identical to the one we wish to scan.

%% Replication 3
To further illustrate the difference between replication and
clusterization, consider an experiment measuring the effect of
rainfall on grass growth \cite{fay2000rainfall}. The experiment
involved four rainfall manipulation "treatments" (i.e.~simulating
different timing and quantity of rainfall), each replicated three
times over a different plot of land. Indeed, it seems reasonable for
researchers to replicate the phenomenon they are trying to study. A
clustered design in such an experiment would imply the researchers
should take a repeated measurement \emph{on the same plot}, at the
expense of measuring grass growth in other plots. To conclude our
short discussion on replication vs.~clusterization: these are
fundamentally different concepts and even though replication is quite
intuitive, it is inapplicable to the inverse problems we consider in
this manuscript.

@@

%% Gaussian
One potential cause of clusterization is our choice of prior. Gaussian
priors, coupled with a Gaussian likelihood and a linear forward
problem give rise to a closed form solution for the posterior via
conjugacy. As we show in Section \ref{subsec:lemma_sims}, such
assumptions generically give rise to clusterization. While an
assumption of a Gaussian likelihood is standard, and rooted in the
central limit theorem, an assumption of Gaussian prior is merely a
matter of convenience. Therefore, we advise any practitioner who
encounters clusterization to replace their Gaussian priors with
non-Gaussian priors instead \cite{hosseini2017, hosseini2019}.

%% Linearization
Another potential cause of clusterization is linearization. While in a
real applications the model is not necessarily linear, some authors
consider a linearized version of their model when seeking optimal
designs \cite{fedorov1996, neitzel2019sparse}. Our analysis then shows
that the linearity of the forward problem is also an important
ingredient in giving rise to clustered desgins. Therefore, our advice
to practitioners is to avoid linearization, and find D-optimal designs
via other methods, e.g.~the sampling method of \cite{ryan2003}.

%% Avoid
Given the above discussion, it is our belief that the methods
suggested by other authors to avoid clusterization are merely
overlooking the problem. We do not view clustered designs as
inherently bad on their own. Rather, we suggest that if a clustered
design arises, a practitioner should revisit their modeling
assumptions --- specifically, their choice of prior and model
linearization. If the practitioner is confident in their assumptions,
then clustered designs should be avoided only to the extent
necessitated by the physical measuring apparatus.

%% Convergence
Given that D-optimal designs suffer from clusterization, it is natural
to ask whether D-optimal designs should be pursued at all. Other
authors have established convergence results for decaying measurement
error \cite{knapik2011}, space-filling measurements
\cite{teckentrup2020} and randomly sampled designs
\cite{nickl2023}. Unfortunately, D-optimal designs --- clustered
designs in particular --- do not fall into any of these categories.

Luckily, we can expect convergence from D-optimal designs (including
clustered designs). Convergence holds in the following sense: the
posterior uncertainty ellipsoid in $\hilo$ will contract to zero along
every eigenfunction of $\fwd \prcov \fwd^*$. See Section
\ref{subsub:convergence} for a proof based on Theorem \ref{thm:char}
--- the main theorem of this manuscript.

\subsection{Limitations}\label{subsec:limitations}
The main limitation of this study is that our generic model does not
correspond to any specific real-life problem. Specifically, in its
current form, our model does not allow point evaluations. Thus, while
our model is generic enough to be analytically tractable, one may
argue that our model is too far removed from any real application. To
these claims I would answer that scientists have a long history of
studying models that are bare-bones simplifications of real systems,
e.g.~the Ising model \cite{cipra1987}, the Lorenz system \cite{brin},
the Lotka-Volterra equations \cite{logan2006}, the Carnot engine
\cite{kardar2007} and many others.

%% When clusterization arises, we believe it should serve as a warning
%% signal to practitioners. In the inverse problem of the 1D heat
%% equation, clusterization occurs primarily because Laplacian
%% eigenvectors do not decay in $\Omega$. Consequently, measuring $u(x_1,
%% T)$ at some point $x_1 \in \Omega$ provides information about
%% $u(x_2,T)$ for distant points $x_2 \in \Omega$. Intuitively, this
%% should not happen: for small $T$, the heat distribution at $x_1$
%% should give very little knowledge on the heat distribution at
%% $x_2$. This behavior stems from a well-known property of the heat
%% equation: it allows information to spread instantly across the
%% computational domain \cite{renardy2006PDE}. In reality, heat (and
%% information) propagate at finite speeds. Of course, the known physical
%% barrier for information spread is the speed of light, but we expect
%% heat and information to spread considerably slower. Thus,
%% clusterization may suggest that our physical model itself is at fault.

@@

@@

### 2024-08-25 latex/vanishing.tex
answered refs

\subsection{Answer to Question \ref{q:why}}\label{subsec:why}
Building on Theorem \ref{thm:char}, we can now give a compelling
explanation to the measurement clusterization we observed for the
inverse problem of the heat equation, see Section
\ref{subsub:clusterization1} below. We also suggest a generic
explanation for clusterization, see Section \ref{subsub:clusterization2}.

@@

>>>>

%% Why
\subsection{Answer to Question \ref{q:why}}\label{subsec:why}
Building on Theorem \ref{thm:char}, we can now give a compelling
explanation to the measurement clusterization we observed for the
inverse problem of the heat equation, see Section
\ref{subsub:clusterization1} below. We also suggest a generic
explanation for clusterization, see Section \ref{subsub:clusterization2}.

@@

\subsubsection{Convergence}\label{subsub:convergence}
Below we fulfill our promise from Section \ref{subsub:implications}
and prove that the posterior uncertainty ellipsoid in $\hilo$ will
contract to zero along every eigenfunction of $\fwd \prcov \fwd^*$. In
our proof we ignore potential problems with conducting inference on
function spaces which are not unique to D-optimal designs (see
e.g.~\ref{owhadi2015} for more details).

First, let $\opt_m$ a D-optimal design utilizing $m$
measurements. Then, an immediate consequence of Theorem \ref{thm:char}
is that $\rank \opt_m^*\opt_m = k \to \infty$ as the number of
measurements $m \to \infty$. Moreover, part (5) of Theorem
\ref{thm:char} ensures that the eigenvalues of the posterior
covariance equal
\begin{equation*}
  \theta^{(k)}_i = \left ( \frac{\sum_{j=1}^k \lambda_j^{-1} +
    \sigma^{-2}m}{k} \right )^{-1}, \text{ for $i\leq k$},
\end{equation*}
where we explicitly express the dependence of $k$ in the
superscript. Now note that $\theta^{(k)}_i \leq \lambda_i$ for $i\leq
k$. Specifically, for fixed $k$, we get $\theta^{(k)}_i \leq \lambda_k$
for all $i\leq k$. Since $\lim_{k\to \infty} \lambda_k= 0$, we
conclude that for all $i$, $\lim_{k\to\infty} \theta^{(k)}_i =
0$. Since $m\to \infty$ implies $k\to \infty$ we have
$\lim_{m\to\infty} \theta^{(k)}_i = 0$ for all $i$. Therefore,
posterior uncertainty decays to zero for all eigenvectors.

### 2024-08-25 revision/rev1.tex
answered refs

\begin{quote}
  Given the above discussion, it is our belief that the methods
  suggested by other authors to avoid clusterization are merely
  overlooking the problem. We do not view clustered designs as
  inherently bad on their own. Rather, we suggest that if a clustered
  design arises, then a practitioner should revisit their probabilistic
  and physical modeling assumptions. If the practitioner is confident in
  their assumptions, then clustered designs should be avoided only to
  the extent necessitated by the physical measuring apparatus.
\end{quote}

@@

\AR In my view, this is a result of the unlocalized structure of
Laplacian eigenvectors. I believe this is more of a problem with our
physical models and choice of priors. See discussion below reproduced
from the revised manuscript:

\begin{quote}
  When clusterization arises, we believe it should serve as a warning
  signal to practitioners. In the inverse problem of the 1D heat
  equation, clusterization occurs primarily because Laplacian
  eigenvectors do not decay in $\Omega$. Consequently, measuring
  $u(x_1, T)$ at some point $x_1 \in \Omega$ provides information
  about $u(x_2,T)$ for distant points $x_2 \in \Omega$. Intuitively,
  this should not happen: for small $T$, the heat distribution at
  $x_1$ should give very little knowledge on the heat distribution at
  $x_2$. This behavior stems from a well-known property of the heat
  equation: it allows information to spread instantly across the
  computational domain \cite{renardy2006PDE}. In reality, heat (and
  information) propagate at finite speeds. Of course, the known
  physical barrier for information spread is the speed of light, but
  we expect heat and information to spread considerably slower. Thus,
  clusterization may suggest that our physical model itself is at
  fault.

  Our choice of prior is perhaps an even greater contributor to
  clusterization. Our choice of Gaussian prior similarly implies
  information is shared between distant locations in $\Omega$. Thus,
  we suggest refraining from choosing Gaussian priors with inverse
  Laplacian covariance operators. Rather, non-Gaussian priors should
  be employed instead \cite{hosseini2017, hosseini2019}. So,
  clusterization may suggest we give more attention to the choice of
  prior.
\end{quote}

@@

\AR I think this is the same misunderstanding I commented on above;
For any nonzero observation error $\sigma^2 > 0$, repeated
measurements will indeed reduce the signal-to-noise ratio of the
measurement and result in an increase of the design criterion. I
briefly comment on this subject in the paragraph below:

\begin{quote}
  Clusterization should not be confused with replication. Replication
  requires that the experimentalist executes multiple trials under
  circumstances that are \emph{nominally identical} \cite[Section
    1.2.4]{morris2011}. Replication is commonly viewed as a beneficial
  and even necessary aspect of optimal experimental design
  \cite{fisher1949design, morris2011,
    schafer2001replication}. Similarly to a design implementing
  replication, a clustered design reduces the signal-to-noise ratio of
  the repeated measurements \cite{telford2007brief}. The difference is
  that a clustered design takes repeated measurements at the expense
  of other quantities not measured at all.
\end{quote}

>>>>

\begin{quote} % Avoid
Given the above discussion, it is our belief that the methods
suggested by other authors to avoid clusterization are merely
overlooking the problem. We do not view clustered designs as
inherently bad on their own. Rather, we suggest that if a clustered
design arises, a practitioner should revisit their modeling
assumptions --- specifically, their choice of prior and model
linearization. If the practitioner is confident in their assumptions,
then clustered designs should be avoided only to the extent
necessitated by the physical measuring apparatus.
\end{quote}

@@

\AR In my view this is a generic phenomenon that arises from a choice
of a linear(ized) model and a Gaussian prior:
\begin{quote}% Gaussian
  One potential cause of clusterization is our choice of prior. Gaussian
priors, coupled with a Gaussian likelihood and a linear forward
problem give rise to a closed form solution for the posterior via
conjugacy. As we show in Section \ref{subsec:lemma_sims}, such
assumptions generically give rise to clusterization. While an
assumption of a Gaussian likelihood is standard, and rooted in the
central limit theorem, an assumption of Gaussian prior is merely a
matter of convenience. Therefore, we advise any practitioner who
encounters clusterization to replace their Gaussian priors with
non-Gaussian priors instead \cite{hosseini2017, hosseini2019}.
\newline %% Linearization
Another potential cause of clusterization is linearization. While in a
real applications the model is not necessarily linear, some authors
consider a linearized version of their model when seeking optimal
designs \cite{fedorov1996, neitzel2019sparse}. Our analysis then shows
that the linearity of the forward problem is also an important
ingredient in giving rise to clustered desgins. Therefore, our advice
to practitioners is to avoid linearization, and find D-optimal designs
via other methods, e.g.~the sampling method of \cite{ryan2003}.
\end{quote}

@@

\AR For any nonzero observation error $\sigma^2 > 0$, repeated
measurements will indeed reduce the signal-to-noise ratio of the
measurement and result in an increase of the design criterion. I
discuss this issue below:

\begin{quote} %% Replication 1
  Clusterization should not be confused with replication. Replication
requires that the experimentalist executes multiple trials under
circumstances that are \emph{nominally identical} \cite[Section
  1.2.4]{morris2011}. Replication is commonly viewed as a beneficial
and even necessary aspect of optimal experimental design
\cite{fisher1949design, morris2011, schafer2001replication}. Similarly
to a design implementing replication, a clustered design reduces the
signal-to-noise ratio of the repeated measurements
\cite{telford2007brief}. The difference is that a clustered design
takes repeated measurements at the expense of other quantities not
measured at all.
\newline %% Replication 2
For example, \cite{fisher1949design}, suggested repeating his famous
milk and tea experiment in order "to be able to demonstrate the
predominance of correct classifications in spite of occasional
errors". Unfortunately, in the experiments we consider, replication is
impossible. For example, in the MRI problem, we cannot generate an
individual nominally identical to the one we wish to scan.
\newline %% Replication 3
To further illustrate the difference between replication and
clusterization, consider an experiment measuring the effect of
rainfall on grass growth \cite{fay2000rainfall}. The experiment
involved four rainfall manipulation "treatments" (i.e.~simulating
different timing and quantity of rainfall), each replicated three
times over a different plot of land. Indeed, it seems reasonable for
researchers to replicate the phenomenon they are trying to study. A
clustered design in such an experiment would imply the researchers
should take a repeated measurement \emph{on the same plot}, at the
expense of measuring grass growth in other plots. To conclude our
short discussion on replication vs.~clusterization: these are
fundamentally different concepts and even though replication is quite
intuitive, it is inapplicable to the inverse problems we consider in
this manuscript.
\end{quote}
